Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-01-19 paper.tex
Minor language fixes

\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup (\Phi^+ \setminus (\Sigma^+_r \cap \Sigma^+_s))$ one has $X_\alpha(R) \subseteq N_{rs}$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence a fortiori it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\U(\Sigma_r^- \cap \Sigma_s^-, I)$ is normalized by $\E(\Delta_{\{r, s\}}, R)$, we obtain the assertion of the lemma for $\alpha \in \Delta_{r, s}$.

Now if $\alpha$ lies in $\Phi^+ \setminus (\Sigma_r^+ \cap \Sigma_s^+)$, then it is contained either in $\Delta_r$ or in $\Delta_s$.
Assume, for example, the latter. By the Levi decomposition we have $\U(\Sigma_r^- \cap \Sigma_s^-, I)^{X_\alpha(R)} \subseteq \U(\Sigma_s^-, I) \subseteq A_{rs}$,
which again implies the assertion of the lemma.
\end{proof}

@@

Denote by $L_{rs}$ the set consisting of all elements $g\in \E(\Phi, R, I)$ such that $g \cdot A_{rs} \subseteq A_{rs}$.
It is easy to see that $L_{rs}$ contains $\EP_r(R, I)$ and is normalized by $N_{rs}$, i.\,e. ${L_{rs}}^{N\ssub{rs}}\subseteq L_{rs}$. Indeed, for $g\in N_{rs}$, $h\in L_{rs}$ one has
\begin{equation}\label{rel:NnormL} h^g \cdot A_{rs} = g^{-1} \cdot h \cdot g \cdot A_{rs} \subseteq g^{-1} \cdot h \cdot A_{rs} \cdot g \subseteq {A_{rs}}^g \subseteq A_{rs}.\end{equation}

@@

The next goal is to demonstrate the inclusion $\U(\Sigma_s^+, R) \subseteq N_{rs}$ (which implies $\EP_s(R) \subseteq N_{rs}$).
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^+_s$ have the same $s$-shape and length, hence the assertion follows from \cref{item-trans2}, \eqref{rel:NnormL} and the fact that we already have the inclusion $X_{\alpha_s}(R) \subseteq N_{rs}$.
 \item \textit{Case $\Phi = \rB_\ell$.} Using the argument from the previous case, we get that $N_{rs}$ contains the subgroup generated by the root subgroups $X_\alpha(R)$, where $\alpha$ varies over long roots of $\Sigma_s^+$. 
  It remains to prove the inclusion $X_{1, 0}(R) \subset N_{rs}$.
  By Chevalley commutator formula, we get
\begin{equation*} \label{rel:CCF-specBC}
x_{1, 0}(ab) = [x_{1, 2}(a), x_{2, 0}(b)] \cdot x_{1, -2}(-a b^2).
\end{equation*}
  Since $X_{2, 0}(R)$ and $X_{1, -2}(R)$ are contained in $N_{rs}$, the inclusion $X_{1, 0}(R) \subseteq N_{rs}$ follows.
 \item \textit{Case $\Phi = \rC_\ell, \rF_4$.}
 We settle these remaining cases in \cref{lemma:DVST}, where we invoke the stability conditions one more time. 
 In fact, we prove a stronger result, namely that $N_{rs}$ contains $X_{-\alpha\ssub{s}}(R)$ and, consequently, $N_{rs} = \E(\Phi, R)$.
\end{enumerate}

In view of \cite[Theorem~3.4]{S} to prove the theorem it suffices to show that $L_{rs}$ contains the generating set $\mathcal{Z}(\Sigma_s^-)$ of $\E(\Phi, R, I)$.
From the inclusion $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)}$ and \eqref{rel:NnormL} it follows that we only have to show that $\mathcal{X} \subseteq L_{rs}$.

The last inclusion follows from the same transitivity argument as above and the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
For $\Phi=\rB_\ell$ one has to once again apply the Chevalley commutator formula:
\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}

@@

\section{Applications} \label{sec:applications}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
Recall from~\cite{LNS11, V13} that a Chevalley group over a field $\G(\Phi, F)$ can be presented as a product of at most $3|\Phi^+|$ of its subgroups $\SL(2, F)$.
As an easy corollary of relative Bass--Kolster decomposition we show that similar decomposition exists in the relative situation and for even larger class of rings.
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system of rank $\ell$ satisfying one the following 4 conditions:
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \ell - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\ell.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ of its subgroups, 
 where each subgroup is an isomorphic copy of $\SL(2, R, I)$.
\end{thm}

>>>>

\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup (\Phi^+ \setminus (\Sigma^+_r \cap \Sigma^+_s))$ one has $X_\alpha(R) \subseteq N_{rs}$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\U(\Sigma_r^- \cap \Sigma_s^-, I)$ is normalized by $\E(\Delta_{\{r, s\}}, R)$, we obtain the result for $\alpha \in \Delta_{r, s}$.

Now if $\alpha$ lies in $\Phi^+ \setminus (\Sigma_r^+ \cap \Sigma_s^+)$, it is contained either in $\Delta_r$ or $\Delta_s$.
Assume, for example, the latter. By the Levi decomposition we have $\U(\Sigma_r^- \cap \Sigma_s^-, I)^{X_\alpha(R)} \subseteq \U(\Sigma_s^-, I) \subseteq A_{rs}$,
which again implies the assertion of the lemma.
\end{proof}

@@

Denote by $L_{rs}$ the set consisting of elements $g\in \E(\Phi, R, I)$ such that $g \cdot A_{rs} \subseteq A_{rs}$.
It is easy to see that $L_{rs}$ contains $\EP_r(R, I)$ and is normalized by $N_{rs}$, i.\,e. ${L_{rs}}^{N\ssub{rs}}\subseteq L_{rs}$. Indeed, for $g\in N_{rs}$, $h\in L_{rs}$ one has
\begin{equation}\label{rel:NnormL} h^g \cdot A_{rs} = g^{-1} \cdot h \cdot g \cdot A_{rs} \subseteq g^{-1} \cdot h \cdot A_{rs} \cdot g \subseteq {A_{rs}}^g \subseteq A_{rs}.\end{equation}

@@

Our next goal is to demonstrate the inclusion $\U(\Sigma_s^+, R) \subseteq N_{rs}$ (which would imply $\EP_s(R) \subseteq N_{rs}$).
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^+_s$ have the same $s$-shape and length, hence the assertion follows from \cref{item-trans2}, \eqref{rel:NnormL} and the fact that we already have the inclusion $X_{\alpha_s}(R) \subseteq N_{rs}$.
 \item \textit{Case $\Phi = \rB_\ell$.} Using the argument from the previous case, we get that $N_{rs}$ contains the subgroup generated by the root subgroups $X_\alpha(R)$, where $\alpha$ varies over the long roots of $\Sigma_s^+$. 
  It remains to prove the inclusion $X_{1, 0}(R) \subset N_{rs}$.
  From Chevalley commutator formula we obtain
\begin{equation*} \label{rel:CCF-specBC}
x_{1, 0}(ab) = [x_{1, 2}(a), x_{2, 0}(b)] \cdot x_{1, -2}(-a b^2).
\end{equation*}
  Since $X_{2, 0}(R)$, $X_{1, -2}(R)$ and $X_{1,2}(R)$ are contained in $N_{rs}$, the inclusion $X_{1, 0}(R) \subseteq N_{rs}$ follows.
 \item \textit{Case $\Phi = \rC_\ell, \rF_4$.}
 We settle these remaining cases in \cref{lemma:DVST}, where we invoke the stability conditions one more time. 
 In fact, we prove a stronger assertion, namely that $N_{rs}$ contains $X_{-\alpha\ssub{s}}(R)$ and, consequently, that $N_{rs} = \E(\Phi, R)$.
\end{enumerate}

In view of \cite[Theorem~3.4]{S} to prove the theorem it suffices to show that $L_{rs}$ contains the generating set $\mathcal{Z}(\Sigma_s^-)$ of $\E(\Phi, R, I)$.
From the inclusion $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)}$ and \eqref{rel:NnormL} it follows that we only need to verify that $\mathcal{X} \subseteq L_{rs}$.
In turn, the last inclusion follows from the same transitivity argument as above if we use the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
For $\Phi=\rB_\ell$ we need to apply Chevalley commutator formula once again:
\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}

@@

\section{Applications} \label{sec:applications}
\subsection{Subsystem factorizations}\label{sec:subsysfact}
Recall from~\cite{LNS11, V13} that a Chevalley group over a field $\G(\Phi, F)$ can be presented as a product of at most $3|\Phi^+|$ of its subgroups $\SL(2, F)$.
As an easy corollary of the relative Bass--Kolster decomposition we get that a similar factorization exists for relative groups over a rather general class of commutative rings.
\begin{thm}\label{thm:SL2width}
Let $I\trianglelefteq R$ be an ideal, $\Phi$ be an irreducible classical root system of rank $\ell$ satisfying one the following 4 conditions:
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right| - \ell - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right| - 2; \\
\Phi=\rB_\ell, \rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right| - 3\ell.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi, R, I)$ can be presented as a product of at most $N$ of its subgroups, 
 where each subgroup is an isomorphic copy of $\SL(2, R, I)$.
\end{thm}
